Location and Scale-Invariant Power Transformations for Transforming Data to Normality

### 2025-10-28 manuscript_rmarkdown/manuscript.tex
Update discussion and caption for figure 5.

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  a4paper,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{booktabs}
\newtheorem*{definition}{Definition}
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator{\sgn}{sgn}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Location and Scale-Invariant Power Transformations for Transforming Data to Normality},
  pdfauthor={Alex Zwanenburg, Steffen Löck},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Location and Scale-Invariant Power Transformations for
Transforming Data to Normality}
\author{Alex Zwanenburg, Steffen Löck}
\date{2025-04-11}

@@

Box-Cox and Yeo-Johnson transformations are modified by introducing
shift parameter \(x_0\) and scale parameter \(s\) into equations
\ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}. The
location- and scale-invariant Box-Cox transformation of a feature value
\(x_i\) of feature \(\mathbf{X}\) under transformation parameter
\(\lambda\), shift parameter \(x_0\) and scale parameter \(s\) is then
defined as:

@@

\begin{itemize}
\item
  The weighting function uses empirical probabilities of the
  distribution of the original feature \(\mathbf{X}\). After sorting
  \(\mathbf{X}\) in ascending order, probabilities are determined as
  \(p_i = \frac{i - 1/3}{n + 1/3}\), with \(i = 1, 2, \ldots n\), with
  \(n\) the number of instances of feature \(\mathbf{X}\). Then
  \(\dot{x}_i = p^{*}_i=2 \left( p_i - 0.5\right)\), so that argument is
  zero-centered.
\item
  The weighting function uses the z-score of the transformed feature
  \(\phi^{\lambda, x_0, s} (\mathbf{X})\). After Raymaekers and
  Rousseeuw (2024),
  \(z_i = \frac{\phi^{\lambda, x_0, s}(x_i) - \mu_M}{\sigma_M}\). Here,
  \(\mu_M\) and \(\sigma_M\) are robust Huber M-estimates of location
  and scale of the transformed feature
  \(\phi^{\lambda, x_0, s} (\mathbf{X})\) (Huber 1981). Then
  \(\dot{x}_i = z_i\).
\item
  After sorting \(\mathbf{X}\) in ascending order, the weighting
  function uses the residual error between the z-score of the
  transformed feature \(\phi^{\lambda, x_0, s} (\mathbf{X})\) and the
  theoretical z-score from a standard normal distribution:
  \(r_i =\left| \left( \phi^{\lambda, x_0, s}(x_i) - \mu_M)\right) / \sigma_M - F^{-1}_{\mathcal{N}}(p_i) \right|\),
  with \(\mu_M\), \(\sigma_M\) and \(p_i\) as defined above. Then
  \(\dot{x}_i = r_i\).
\end{itemize}

@@

We require a quantile function (or an approximation thereof) to draw
random values from an asymmetric generalised normal distribution using
inverse transform sampling. Gijbels et al.~derived the quantile function
\(F_{\alpha}^{-1}(p)\), which incorporates the quantile function of the
symmetric generalised normal distribution derived by Griffin (Griffin
2018):

@@

\subsection{Central normality test}\label{central-normality-test}

Power transformations aim to transform features to a normal
distribution. However, this may not always be successful or possible.
Deviations from normality can be detected by normality tests, such as
the Shapiro-Wilk test (Shapiro and Wilk 1965). In practice, normality
tests may be too stringent with large sample sizes, outliers, or both.
Here we develop a test for central normality.

@@

\begin{definition}[Central normality of a sequence]

The central portion of a sequence is normally distributed if the sum of residual
errors is equal to zero:
$\sum_{r_j \in \mathbf{R}_{\text{c}}} r_j = 0$

@@

In practice, a finite sequence sampled from a normal distribution
\(\mathcal{N}(\mu, \sigma)\) will have a non-zero sum of residual
errors.

\begin{definition}[Central normality test]
The null-hypothesis is defined as: $\mathcal{H}_0: \sum_{r_j \in \mathbf{R}_{\text{c}}}r_j = 0$

The alternative hypothesis is defined as: $\mathcal{H}_1: \sum_{r_j \in \mathbf{R}_{\text{c}}}r_j > 0$
\end{definition}

The null-hypothesis \(\mathcal{H}_0\) is that the central portion of
sequence \(\mathbf{X}\) is normally distributed, with the alternative
hypothesis \(\mathcal{H}_1\) that it is not normally distributed. To
test this hypothesis, a test statistic is computed and compared against
a critical value.

\begin{definition}[Central normality test statistic]
Let $m$ be the number of elements of $\mathbf{R}_{\text{c}}$

@@

The null-hypothesis should be rejected at significance level \(\alpha\)
if \(\tau_{n, \kappa} \geq \tau_{\alpha, n, \kappa, \text{critical}}\).

The central portion of the data needs to be defined and the Type 1 error
rates determined to provide critical test statistics. We will do so in
the \hyperref[simulation]{Simulation} section.

@@

We used simulated data to assess invariance to location and scale of the
proposed power transformations, to develop the empirical central
normality test, and to determine weighting for robust transformations.
The \(\lambda\) parameter for conventional power transformations (Eqn.
\ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}), as well
as \(\lambda\), \(x_0\) and \(s\) parameters for location- and
scale-invariant power transformations (Eqn. \ref{eqn:box-cox-invariant}
and \ref{eqn:yeo-johnson-invariant}) were estimated using the BOBYQA
algorithm for derivative-free bound constraint optimisation (Powell
2009) through maximum likelihood estimation. The required algorithms
were implemented in the \texttt{power.transform} R software package
(Zwanenburg and Löck 2024b) (version 1.0.1). Of note, the
\texttt{power.transform} package shifts feature values into the positive
domain if negative or zero values are present for Box-Cox power
transformations.

@@

\subsection{Central normality and empirical central normality
test}\label{central-normality-and-empirical-central-normality-test}

To develop a test for central normality we need to consider two
parameters: the central portion \(\kappa\) as a fixed parameter, and
test statistic \(\tau_{\text{ecn}}\). We will first define the central
portion \(\kappa\).

For each number of samples
\(n \in \left\{\lfloor 10^\nu \rfloor | \nu \in \left\{0.7500, 0.8125, \ldots, 4.0000 \right \} \right\}\)
we randomly drew \(m_d = 30000\) ordered sequences \(\mathbf{X}\) from
\(\mathcal{N}(0,1)\). This dataset was used to compute the critical test
statistic values for the central normality test. Additionally, we added
\(10 \%\) outliers by randomly replacing elements of each ordered
sequence. The dataset with outliers was used to compute the critical
test statistic values for the empirical variant of the central normality
test. For each sequence we then computed \(\tau_{n, \kappa}\) for
\(\kappa \in \left\{0.60, 0.70, 0.80, 0.90, 0.95, 1.00\right\}\).

Figure \ref{fig:empirical-central-normality-kappa} shows
\(\tau_{\alpha = 0.05, n, \kappa}\) as a function of \(n\) for different
values of \(\kappa\). With decreasing \(\kappa\), the test statistic
curve decreases. The curves for \(\kappa = 0.60\), \(\kappa = 0.70\) and
\(\kappa = 0.80\) are similar, whereas curves for \(\kappa \geq 0.90\)
are affected by outliers, as can be observed by comparing curves of the
central normality test with those of the empirical variant.

Since the (empirical) central normality tests assesses whether the
central part of a sequence is normally distributed, we used
\(\kappa = 0.80\). Critical statistic values for central normality and
empirical central normality tests are shown are shown in Table
\ref{tab:central-normality-critical-statistic} and Table
\ref{tab:empirical-central-normality-critical-statistic}, respectively.

@@

{\centering \includegraphics{manuscript_files/figure-latex/empirical-central-normality-kappa-1}

@@

\caption{Critical test statistic $\tau_{\alpha = 0.05, n, \kappa}$ of the (empirical) central normality test as function of $n$ for several values of central portion $\kappa$. The critical test statistics for central normality test are determined using fully normal data, whereas the statistics for the empirical variant are determined using centrally normal data, i.e. with fully normal data where 10\% of elements are replaced by outliers. emp: empirical}\label{fig:empirical-central-normality-kappa}
\end{figure}

To assess type I error rates for the (empirical) central normality test
and compare these to the Shapiro-Wilk test for normality, we randomly
drew another 1000 sequences from a standard normal distribution
(\(\mathcal{N}(0,1)\) for each
\(n \in \left\{\lfloor 10^\nu \rfloor | \nu \in \left\{0.7500, 0.8125, \ldots, 3.0000 \right \} \right\}\).
A second dataset was created by replacing elements with randomly drawn
outliers, as described above. For each sequence, the p-value for the
respective test was used to reject the null hypothesis that the sequence
is (centrally) normal. The null hypothesis was rejected if
\(p \leq 0.05\). The type I error rate was determined by computing the
fraction of centrally normal sequences rejected this way. Figure
\ref{fig:empirical-central-normality-error_rate} shows that the
empirical central normality test had the expected type I error rate for
centrally normal data (with tails of the distribution not being normally
distributed), whereas the central normality test and Shapiro-Wilk test
overestimated type I error rates.

{\centering \includegraphics{manuscript_files/figure-latex/empirical-central-normality-error_rate-1}

\caption{Observed type I error rates for (central) normality tests in the presence of data derived from normal distributions with and without outliers. The empirical central normality (ECN) test uses $\kappa = 0.80$. CN: central normality}\label{fig:empirical-central-normality-error_rate}
\end{figure}

In Figure \ref{fig:empirical-central-normality-examples} we apply the
empirical central normality test to assess central normality of features
that are composed of a mixture of samples drawn from two normal
distributions (\(n = 100\) each). With increased separation of the
underlying normal distributions, the probability of the feature being
centrally normal decreases, as expected.

@@

\caption{Bi-modal distributions and empirical central normality test results. The feature (black) is a mixture of two identical sample sets (blue and orange, $n = 100$) drawn from normal distributions that are offset by a distance $d$. We use the empirical centrally normality test to compute the probability for the hypothesis that the distribution is centrally normal. As may be observed, with increasing offset $d$ the probability that the feature is centrally normal decreases. Quantile-quantile plots are drawn below each distribution.}\label{fig:empirical-central-normality-examples}
\end{figure}

@@

To determine the weighting function parameters for each of the nine
combinations, \(m_d=500\) sequences \(\mathbf{X}_i\)
(\(i \in \{1, 2, \ldots, m_d\}\)) were randomly drawn from randomly
parametrised asymmetric generalised normal distributions. Each
distribution was parametrised with a random skewness parameter
\(\alpha \sim U\left(0.01, 0.99\right)\) and shape parameter
\(\beta \sim U\left(1.00, 5.00 \right)\). Location and scale parameters
were set as \(\mu = 0\) and \(\sigma = 1\), respectively. To form each
sequence \(\mathbf{X}_i\), \(n = \lceil 10^\gamma \rceil\) instances
were then randomly drawn, with
\(\gamma \sim U\left(\log_{10}50, 3\right)\), resulting in a set of
sequences with between \(50\) and \(1000\) elements each.

Outlier values were then drawn to randomly replace 10 percent of the
elements of \(\mathbf{X}_i\). Outlier values were set according to John
W. Tukey (1977), as follows. Let \(x^{*} \sim U\left(-2, 2\right)\).
Then the corresponding outlier value was:

\begin{equation}
x_{out} =
\begin{cases}
Q_1 - \left(1.5 - x^{*} \right) \text{IQR} & \text{if } x^{*} < 0 \\
Q_3 + \left(1.5 + x^{*} \right) \text{IQR} & \text{if } x^{*} \geq 0
\end{cases}
\end{equation}

\(Q_1\), \(Q_3\) and \(\text{IQR}\) are the first quartile, third
quartile and interquartile range of \(\mathbf{X}_i\), respectively.
Outlier values randomly replaced elements in \(\mathbf{X}_i\).

@@

Three datasets with 10000 sequences each were created to assess
transformation to normality. For each sequence \(\mathbf{X}_i\),
\(n = \lceil 10^\gamma \rceil\) elements were randomly drawn, with
\(\gamma \sim U\left(1, 4\right)\), resulting in sequences with between
\(10\) and \(10000\) elements.

@@

In addition to no power transformation and location- and scale-invariant
power transformations, conventional and Raymaekers and Rousseeuw's
robust adaptation (Raymaekers and Rousseeuw 2024) were assessed. For the
latter two methods, normalisation before standardisation using was
additionally assessed using the following two methods:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  z-standardisation:
  \(x^{\prime}_{i} = \left(x_i - \mu \right) / \sigma\), with \(\mu\)
  and \(\sigma\) the mean and standard deviation of sequence
  \(\mathbf{X}\).
\item
  robust scaling:
  \(x^{\prime}_{i} = \left(x_i - \text{median}\left(\mathbf{X}\right) \right) / \text{IQR}\left(\mathbf{X}\right)\),
  with \(\text{IQR}\) representing the interquartile range of
  \(\mathbf{X}\).
\end{enumerate}

This results in nine types of power transformation. Transformation
parameters were optimised for each sequence. Subsequently the sum of
residual errors and sum of residual errors of the central portion
(\(\kappa = 0.80\)) were computed for each sequence after
transformation. Then, each method was ranked according to the sum of
residual errors (data without outliers) or the sum of residual errors of
the central portion (data with outliers). The average rank of each
transformation method was computed over all 10000 sequences in each
dataset. Average ranks for Yeo-Johnson transformations are shown in
Table \ref{tab:comparison_methods_simulation_yeo_johnson}. Location and
shift-invariant Yeo-Johnson transformation ranked best for all datasets
without outliers. The robust variant ranked best in dirty and shifted
datasets with outliers, but not in the clean dataset. Results for
Box-Cox transformations are shown in Table
\ref{tab:comparison_methods_simulations_box_cox}.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\documentclass[
  a4paper,
]{article}
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\usepackage[]{natbib}
\bibliographystyle{plainnat}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{booktabs}
\newtheorem*{definition}{Definition}
\newtheorem*{corollary}[Corollary]
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator{\sgn}{sgn}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Location and Scale-Invariant Power Transformations for Transforming Data to Normality},
  pdfauthor={Alex Zwanenburg, Steffen Löck},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Location and Scale-Invariant Power Transformations for
Transforming Data to Normality}
\author{Alex Zwanenburg, Steffen Löck}
\date{2025-10-28}

@@

\citet{Box1964-mz} originally included a shift parameter in their
transformation, but for their formal analysis set it to \(0\), whereas
\citet{Yeo2000-vw} did not include a shift or scale parameter. Here,
Box-Cox and Yeo-Johnson transformations are modified by introducing
shift parameter \(x_0\) and scale parameter \(s\) into equations
\ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}. The
location- and scale-invariant Box-Cox transformation of a feature value
\(x_i\) of feature \(\mathbf{X}\) under transformation parameter
\(\lambda\), shift parameter \(x_0\) and scale parameter \(s\) is then
defined as:

@@

\begin{itemize}
\item
  The weighting function uses empirical probabilities of the
  distribution of the original feature \(\mathbf{X}\). After sorting
  \(\mathbf{X}\) in ascending order, probabilities are determined as
  \(p_i = \frac{i - 1/3}{n + 1/3}\), with \(i = 1, 2, \ldots n\), with
  \(n\) the number of instances of feature \(\mathbf{X}\). Then
  \(\dot{x}_i = p^{*}_i=2 \left( p_i - 0.5\right)\), so that argument is
  zero-centered.
\item
  The weighting function uses the z-score of the transformed feature
  \(\phi^{\lambda, x_0, s} (\mathbf{X})\). After
  \citet{Raymaekers2024-zf},
  \(z_i = \frac{\phi^{\lambda, x_0, s}(x_i) - \mu_M}{\sigma_M}\). Here,
  \(\mu_M\) and \(\sigma_M\) are robust Huber M-estimates of location
  and scale of the transformed feature
  \(\phi^{\lambda, x_0, s} (\mathbf{X})\) \citep{Huber1981-su}. Then
  \(\dot{x}_i = z_i\).
\item
  After sorting \(\mathbf{X}\) in ascending order, the weighting
  function uses the residual error between the z-score of the
  transformed feature \(\phi^{\lambda, x_0, s} (\mathbf{X})\) and the
  theoretical z-score from a standard normal distribution:
  \(r_i =\left| \left( \phi^{\lambda, x_0, s}(x_i) - \mu_M)\right) / \sigma_M - F^{-1}_{\mathcal{N}}(p_i) \right|\),
  with \(\mu_M\), \(\sigma_M\) and \(p_i\) as defined above. Then
  \(\dot{x}_i = r_i\).
\end{itemize}

@@

We require a quantile function (or an approximation thereof) to draw
random values from an asymmetric generalised normal distribution using
inverse transform sampling. Gijbels et al.~derived the quantile function
\(F_{\alpha}^{-1}(p)\), which incorporates the quantile function of the
symmetric generalised normal distribution derived by Griffin
\citep{Griffin2018-bf}:

@@

\subsection{Empirical central normality
test}\label{empirical-central-normality-test}

Power transformations aim to transform features to a normal
distribution. However, this may not always be successful or possible.
Deviations from normality can be detected by normality tests, such as
the Shapiro-Wilk test \citep{Shapiro1965-zd}. In practice, normality
tests may be too stringent with large sample sizes, outliers, or both.
Here we develop an empirical test for central normality.

@@

\begin{definition}[Central normal distribution]

A central normal distribution is any distribution whose quantile function is 
defined by $F^{-1}_{\mathcal{N}}$ for $\frac{1-\kappa}{2} \leq  p \leq \frac{1 + \kappa}{2}$,
with $\mathcal{N}$ any normal distribution.

@@

\begin{corollary}

For $\kappa = 1.0$, the central normal distribution is a normal distribution.

\end{corollary}

\begin{corollary}

By the Glivenko-Cantelli theorem, if a sequence $\mathbf{X}$ was sampled from a
central normal distribution, then almost surely
$\lim_{n \rightarrow \infty} \sum_{r_j \in \mathbf{R}_{\text{c}}} r_j = 0$,

\end{corollary}

\begin{definition}[Central normality test]
The null-hypothesis $\mathcal{H}_0$ is that the sequence $\mathbf{X}$ was sampled
from a population that is distributed according to a central normal distribution.

The alternative hypothesis is that the sequence $\mathbf{X}$ was sampled from a
population that is not distributed according to a central normal distribution.
\end{definition}

To test this hypothesis, a test statistic is computed and compared
against a critical value.

\begin{definition}[Central normality test statistic]
Let $m$ be the number of elements of $\mathbf{R}_{\text{c}}$.

@@

The null-hypothesis should be rejected at significance level \(\alpha\)
if \(\tau_{n, \kappa} \geq \tau_{\alpha, n, \kappa, \text{critical}}\).
Critical test statistic values are determined in the
\hyperref[simulation-ecn-test]{Simulation} section.

@@

We used simulated data to assess invariance to location and scale of the
proposed power transformations, to develop the empirical central
normality test, and to determine weighting for robust transformations.
The \(\lambda\) parameter for conventional power transformations (Eqn.
\ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}), as well
as \(\lambda\), \(x_0\) and \(s\) parameters for location- and
scale-invariant power transformations (Eqn. \ref{eqn:box-cox-invariant}
and \ref{eqn:yeo-johnson-invariant}) were estimated using the BOBYQA
algorithm for derivative-free bound constraint optimisation
\citep{Powell2009-zb} through maximum likelihood estimation. The
required algorithms were implemented in the \texttt{power.transform} R
software package \citep{Zwanenburg2024-kq} (version 1.0.1). Of note, the
\texttt{power.transform} package shifts feature values into the positive
domain if negative or zero values are present for Box-Cox power
transformations.

@@

\subsection{Empirical central normality test}\label{simulation-ecn-test}

The critical test statistic for the empirical central normality test
depends on the significance level \(\alpha\), sequence length \(n\) and
central portion \(\kappa\). Critical test statistic values were set
through simulation, as described below.

For each number of samples
\(n \in \left\{\lfloor 10^\nu \rfloor | \nu \in \left\{0.7500, 0.8125, \ldots, 4.0000 \right \} \right\}\)
we randomly sampled \(m_d = 30000\) ordered sequences \(\mathbf{X}\)
from \(\mathcal{N}(0,1)\). For each sequence we then computed
\(\tau_{\kappa}\) for
\(\kappa \in \left\{0.50, 0.60, 0.70, 0.80, 0.90, 0.95, 1.00\right\}\).

Figure \ref{fig:empirical-central-normality-combined} shows the critical
test statistic \(\tau_{\alpha, n, \kappa}\) as a function of \(n\) for
different values of \(\kappa\) and \(\alpha\). For larger central
portion \(\kappa\), critical test statistic values increase. Similarly,
critical test statistic values increase for lower significance levels.

@@

{\centering \includegraphics{manuscript_files/figure-latex/empirical-central-normality-combined-1}

@@

\caption{Critical test statistic $\tau_{\alpha = 0.05, n, \kappa}$ of the empirical central normality test as function of $n$ for several values of central portion $\kappa$ (left) and several values of significance level $\alpha$ (right).}\label{fig:empirical-central-normality-combined}
\end{figure}

To assess type I error rates for the empirical central normality test
and compare these to the Shapiro-Wilk test for normality, we randomly
drew another 10000 sequences from a central normal distribution for each
\(n \in \left\{\lfloor 10^\nu \rfloor | \nu \in \left\{0.7500, 0.8125, \ldots, 3.0000 \right \} \right\}\).

Each sequence was created as follows: first \(n\) values are sampled
from \(\mathcal{N}(0, 1)\). Then, the elements outside the central
portion of the sequence
(i.e.~\(\mathbf{X} \setminus \mathbf{X}_{\text{c}}\)) are replaced.
Elements with \(pi < \frac{1-\kappa}{2}\) are replaced with values
sampled from \(\mathcal{U}(-10,\min(\mathbf{X}_{\text{c}}))\), and
elements with \(p_i > \frac{1 + \kappa}{2}\) are replaced with values
sampled from \(\mathcal{U}(\max(\mathbf{X}_{\text{c}}), 10)\).

For each sequence, the p-value for the respective test was used to
reject the null hypothesis that sequences were derived from a population
with a (central) normal distribution. The null hypothesis was rejected
if \(p \leq 0.05\). The type I error rate was determined by computing
the fraction of sequences rejected this way. Figure
\ref{fig:empirical-central-normality-error_rate} shows that when the
central portion \(\kappa\) of the central normal distribution is equal
or greater than the \(\kappa\) parameter for the empirical central
normality test, the tests are accurate. If \(\kappa\) of the empirical
central normality test exceeds that the central portion \(\kappa\) for
the underlying distribution, tests are conservative. The Shapiro-Wilk
test behaves similar to the empirical central normality test with
\(\kappa = 1.0\), which is expected. For \(n < 10\), the empirical
central normality test can be optimistic.

\textbackslash begin\{figure\}

\{\centering \includegraphics{manuscript_files/figure-latex/empirical-central-normality-error_rate-1}

\}

\textbackslash caption\{Observed type I error rates for central
normality tests in the presence of data sampled from central normal
distributions with \$\kappa \in \left{0.6, 0.8, 1.0\right}\}. Three
empirical central normality (ECN) tests were performed, parameterised
with same value for the central portion \(\kappa\).Additional, the
Shapiro-Wilk test was assessed.ECN: empirical central
normality\}\label{fig:empirical-central-normality-error_rate}
\textbackslash end\{figure\}

In Figure \ref{fig:empirical-central-normality-examples} we apply the
empirical central normality test (with \(\kappa = 0.80\)) to assess
central normality of features that are composed of a mixture of samples
drawn from two normal distributions (\(n = 100\) each). With increased
separation of the underlying normal distributions, the test's p-value
decreases, as expected.

@@

\caption{Bi-modal distributions and empirical central normality test results. The feature (black) is a mixture of two identical sample sets (blue and orange, $n = 100$) drawn from normal distributions that are offset by a distance $d$. The empirical central normality ($\kappa = 0.80$) is used to test the hypothesis that the mixture is sampled from a central normal distribution. As may be observed, with increasing offset $d$ the test's p-value decreases. Quantile-quantile plots are drawn below each distribution.}\label{fig:empirical-central-normality-examples}
\end{figure}

We use \(\kappa = 0.80\) for the empirical central normality tests in
the remainder of the manuscript, unless noted otherwise. Critical
statistic values of the empirical central normality test for
\(\kappa = 0.80\) are shown in Table
\ref{tab:empirical-central-normality-critical-statistic}.

@@

To determine the weighting function parameters for each of the nine
combinations, \(m_d=500\) sequences \(\mathbf{X}_i\)
(\(i \in \{1, 2, \ldots, m_d\}\)) were randomly drawn from randomly
parametrised asymmetric generalised normal distributions. Each
distribution was parametrised with a random skewness parameter
\(\alpha \sim U\left(0.01, 0.99\right)\) and shape parameter
\(\beta \sim U\left(1.00, 5.00 \right)\). Location and scale parameters
were set as \(\mu = 0\) and \(\sigma = 1\), respectively. To form each
sequence \(\mathbf{X}_i\), \(n = \lceil 10^\gamma \rceil\) instances
were then randomly drawn, with
\(\gamma \sim U\left(\log_{10}50, 3\right)\), resulting in a set of
sequences with between \(50\) and \(1000\) elements each. Outlier values
were then drawn to randomly replace 10 percent of the elements of
\(\mathbf{X}_i\). These values were set according to
\citet{Tukey1977-xm}, as follows. Let
\(x^{*} \sim U\left(-2, 2\right)\). Then the corresponding outlier value
was:

\begin{equation}
\label{eqn:outlier_rate}
x_{out} =
\begin{cases}
Q_1 - \left(1.5 - x^{*} \right) \text{IQR} & \text{if } x^{*} < 0 \\
Q_3 + \left(1.5 + x^{*} \right) \text{IQR} & \text{if } x^{*} \geq 0
\end{cases}
\end{equation}

Here, \(Q_1\), \(Q_3\) and \(\text{IQR}\) are the first quartile, third
quartile and interquartile range of \(\mathbf{X}_i\), respectively.

@@

Three datasets with 100000 sequences each were created to assess
transformation to normality. For each sequence \(\mathbf{X}_i\),
\(n = \lceil 10^\gamma \rceil\) elements were randomly drawn, with
\(\gamma \sim U\left(1, 4\right)\), resulting in sequences with between
\(10\) and \(10000\) elements.

@@

In addition to no power transformation and location- and scale-invariant
power transformations, conventional and Raymaekers and Rousseeuw's
robust adaptation \citep{Raymaekers2024-zf} were assessed. For the
latter two methods, normalisation before standardisation using was
additionally assessed using the following two methods:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  z-standardisation:
  \(x^{\prime}_{i} = \left(x_i - \mu \right) / \sigma\), with \(\mu\)
  and \(\sigma\) the mean and standard deviation of sequence
  \(\mathbf{X}\).
\item
  robust scaling:
  \(x^{\prime}_{i} = \left(x_i - \text{median}\left(\mathbf{X}\right) \right) / Q_n\left(\mathbf{X}\right)\),
  with \(Q_n\) representing the robust \(Q_n\) scale estimator of
  \(\mathbf{X}\) \citep{Croux1992-zz, Rousseeuw1993-it}.
\end{enumerate}

This results in nine types of power transformation. Transformation
parameters were optimised for each sequence. Subsequently the sum of
residual errors and sum of residual errors of the central portion
(\(\kappa = 0.80\)) were computed for each sequence after
transformation. Then, each method was ranked according to the sum of
residual errors (data without outliers) or the sum of residual errors of
the central portion (data with outliers). The average rank of each
transformation method was computed over all 100000 sequences in each
dataset. Average ranks for Yeo-Johnson transformations are shown in
Table \ref{tab:comparison_methods_simulation_yeo_johnson}. Location and
shift-invariant Yeo-Johnson transformation ranked best for all datasets
without outliers. The robust variant ranked best in dirty and shifted
datasets with outliers, but not in the clean dataset. Results for
Box-Cox transformations are shown in Table
\ref{tab:comparison_methods_simulations_box_cox}.
